Implementation of a minimal branch-decomposition algorithm for planar graphs.

### 2024-06-24 paper/main.tex
Add some examples to prelim

	A \textit{graph} $G$ is a 3-tuple $G = (V, E, \psi)$ consisting of a vertex set $V$, an edge set $E$ and a function $\psi$, where $V \subset \mathbb{N}^+$ and where $E \subset \mathbb{N}^+$ and where $\psi \st E \to \{\{u,v\} \st u,v \in V \land u \neq v \}$.

	\textit{Note that other authors might instead name this an Undirected Labelled Multigraph (not permitting self-loops) (edges having own identity).}

	A graph $G$ is \textit{simple}, if it has no parallel edges; $\neg \exists e_1,e_2 \in E_G$ where $e_1 \neq e_2$ such that $\psi_G(e_1) = \psi_G(e_1)$.

	A \textit{subgraph} $H$ of a graph $G$, is a graph where some vertices and edges might be missing; a graph where $V_H \subseteq V_G$ and where $E_H \subseteq E_G$ and where $\psi_H$ is a subfunction of $\psi_G$; if $e \in E_H$ then $\psi_H(e) = \psi_G(e)$.

	A graph is \textit{connected} if there exists a path connecting any two vertices.

@@

	A graph is \textit{planar} if there exists a planar embedding of the graph.

	A \textit{plane graph} (syn. \textit{planar embedding of the graph}), is a 2D drawing of a planar graph such that no edges are crossing. In other words, for a planar graph $G$ a \textit{plane graph}, is a set of points $P$ and line segments $L$, with a one-to-one correspondence between $V(G)$ and $P$ and with a one-to-one correspondence between $E(G)$ and $L$, such that one of the endpoints of the line segments for the corresponding edges that are incident to some vertex $v$ occupy the same coordinate as the point corresponding to $v$ and such that line segments only pairwise intersect at their endpoints.

@@

		Given an graph $G$ and pair of vertices $u,v \in G_V$ such that $\{u,v\}$ is in the image of $\psi_G$, remove all edges incident to $u$ and $v$ and update every edge $\{v,w\} \in E$ to be $\{u,w\}$.
	\end{definition}

@@

		The dual graph $M^*$ of a plane graph $M$ is a graph where a face $f$ of $M$ is a vertex $f^*$ of $M^*$ and an edge $e$ that separates two faces $f_1$,$f_2$ of $M$ is an edge $e^*$ of $M^*$ incident to $f_1^*$ and $f_2^*$.
	\end{definition}

	\begin{corollary}
		If multiple edges separate $f_1$ and $f_2$ there will be parallel edges incident to $f_1*$ and $f_2*$.
	\end{corollary}

		The medial graph $M(G)$ of a connected plane graph $G$ is a graph with a vertex $e^*$ for each edge $e$ of $G$ and for each face $f$ of $G$, there's an edge $d^*$ incident to a pair of vertices $e_1^*$, $e_2^*$ of $M(G)$ if $e_1$ and $e_2$ are consecutive in $f$.
	\end{definition}

	A \textit{Branch Decomposition} $B$ of a graph $G$ is a tree where every edge of $G$ is a leaf in $B$ and every internal vertex of $B$ has exactly 3 neighbors. $B$ is an unrooted binary tree.

	Removing any edge $e = \{u,v\}$ of $B$ partitions $B$ into 2 trees $B_u$ and $B_v$ and the intersection of the sets of vertices in the leaves of $B_u$ and $B_v$ is called a middle set. Every edge of $B$ has an associated middle set. The maximal cardinality of any middle set among all middle sets of $B$ is the width of $B$.

	There can be many branch decompositions of a graph $G$.

@@

	Without the constraint of planarity on the input graph, then it is NP-complete to determine whether a general graph $G$ has a branch decomposition of width at most $k$, when $G$ and $k$ are both considered as inputs to the problem.

	Let $n = |V(G)|$ and $m = |E(G)|$.

	A \textit{Carving Decomposition}

	A \textit{Minimal Carving Decomposition}

@@

		Input: Given a simple undirected connected planar graph $G$.

@@

		Input: Given a simple undirected connected graph $G$.

@@

	Problem \ref{def:mbdp} is the overarching problem, that the algorithm solves, and can be broken down into many smaller subproblems.

	Considering a graph $G$, you can compute a minimal branch decomposition of $G$ from a minimal carving decomposition of the medial of $G$, by replacing the vertices in leaves of the decomposition with edges, using the mapping between edges and vertices from computing the medial graph. \ref{} Therefore problem \ref{def:mbdp} break down into problems \ref{problem:mcd-to-mbd}, \ref{problem:medial} and \ref{problem:mcd}.

@@

	A quiet subgraph $Q(M, k, e)$, for some graph $M$, some noise level $k$ and some $e \in E(M)$, is a subgraph of $M$ with $V(Q(M, k, e)) = V(M)$ and

	$E(Q(M, k, e)) = \{e_1 \st $ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

	\begin{problem}\label{problem:noisy-subgraph}
		Given a plane graph $M$ that might have parallel edges, an edge $e \in E(M)$, and noise level $k \in \mathbb{N}$, output the quiet subgraph $Q(M, k, e)$.
	\end{problem}

@@

	Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \st e \in E(M) \land C$ is a component of $Q(M, k, e) \}$.

@@

		I assume that the input graph $G$ is a planar graph given as an adjacency list such that the neighborhoods are given in clockwise ordering according to some plane embedding of $G$.

@@

		As the resulting graph is later given as an argument to functions assuming a clockwise or counterclockwise ordering of vertices, the implementation needs to preserve this invariant when contracting.

		As this contraction is a contraction of ALL edges connecting a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the ordering invariant could be difficult, but luckily in this context, it is irrelevant.

@@

>>>>

	A \textit{graph} $G$ consists of a vertex set $V(G)$, and an edge set $\mathbb{E}(G)$ and a function $\phi_G$, where $V(G) \subset \mathbb{N}^+$ and where $\mathbb{E}(G) \subset \mathbb{N}^+$ and where $\phi_G \st \mathbb{E}(G) \to \{\{u,v\} \st u,v \in V(G)\}$.

	\textbf{Note.} Other authors might instead call this definition an undirected labelled pseudograph, with edges having own identity.

	\textbf{Note.} Regarding notation, $V$ and $\mathbb{E}$ are operations on graphs returning the vertex set and edge set respectively.

	A \textit{drawing} of a graph $G$ is a node-link diagram in which the vertices are represented as disks and the edges are represented as line segments or curves in the Euclidean plane.

	Here is a drawing of a graph $G$.

	\begin{center}
		\input{images/unlabeled-graph.tex}
	\end{center}

	Here is a labeled drawing of the same graph $G$ and its function $\phi_G$.

	\begin{center}
		\input{images/graph.tex}
	\end{center}

	Let $E(G)$ return a multiset of all vertex-pairs of $G$; in other words, $E(G) = \{\phi_G(e) \st e \in \mathbb{E}(G)\}$.

	A edge $e$ where $\phi_G(e) = \{u,v\}$, is a \textit{self-loop}, if $u = v$.

	A graph $G$ is \textit{loop-less}, if no edge $e \in \mathbb{E}(G)$ is a self-loop.

	A \textit{multi-graph} is a graph that is loop-less.

	\begin{center}
		\input{images/multigraph.tex}
	\end{center}

	A multi-graph $G$ is \textit{simple}, if it has no parallel edges; in other words, if elements of $E(G)$ are pair-wise distinct.

	\begin{center}
		\input{images/simple-graph.tex}
	\end{center}

	A \textit{subgraph} $H$ of a graph $G$, is a graph where some vertices and edges might be missing; in other words, is a graph where $V(H) \subseteq V(G)$ and where $\mathbb{E}(H) \subseteq \mathbb{E}(G)$ and where $\forall e \in \mathbb{E}(H), \phi_H(e) = \phi_G(e)$.

	\begin{center}
		\input{images/subgraph.tex}
	\end{center}

	For $A \subseteq V(G)$, we denote by $G[A]$ the subgraph induced by the subset of vertices $A$; in other words, $G[A]$ is the subgraph where $V(G[A]) = A$ and where $\mathbb{E}(G[A]) = \{ e \st e \in \mathbb{E}(G) \land |\phi_G(e) \cap A| = 2\}$ and where $\forall e \in E(G[A]), \phi_{G[A]}(e) = \phi_{G}(e)$.

	\begin{center}
		\input{images/induced-subgraph.tex}
	\end{center}

	A vertex $v \in V(G)$ and an edge $e \in \mathbb{E}(G)$ are \textit{incident} to each other, if $v \in \phi_G(e)$. Furthermore, two distinct edges $e_1,e_2 \in \mathbb{E}(G)$ are incident to each other, if $\phi_G(e_1) \cap \phi_G(e_2) \neq \emptyset$.

	The \textit{degree} of a vertex $v$, denoted $\deg(v)$, is the number of times that an edge is incident to $v$. A self-loop is incident to the same vertex twice.

	\begin{center}
		\input{images/degree.tex}
	\end{center}

	The \textit{maximum degree} of a graph $G$, denoted $\Delta(G)$, is the maximal degree of any vertex of $G$.

	A \textit{walk} of a graph $G$ is a list of edges, such that consecutive edges in the list are incident to each other.

	An \textit{$s,t$-walk} is a walk, such that the first edge is incident to the vertex $s$ and such that the last edge is incident to the vertex $t$.

	The \textit{length} of a walk is the number of edges in the walk.

	An $s,t$-walk is \textit{closed}, if $s=t$.

	A \textit{path} of a graph $G$, is a walk such that only consecutive edges of the walk are incident.

	A \textit{cycle} of a graph $G$, is a path but where the first and the last edge are incident to each other.

	A graph $G$ is \textit{connected} if there exists a $s,t$-walk for every pair of distinct vertices $s,t \in V(G)$.

@@

	A \textit{bijection} is a relation between two sets such that each element of either set is paired with exactly one element of the other set.

	A \textit{plane graph} is a drawing of a graph, such that no edges are crossing.

	A graph $G$ is \textit{planar}, if there exists a \textit{plane graph} of $G$.

@@

		A contraction is a function that given a multi-graph $G$ and pair of vertices $u,v \in V(G)$ such that $\{u,v\} \in E(G)$, then for all edges $e \in \mathbb{E}(G)$ if $\phi_G(e) = \{u,v\}$ then removes $e$ else if $\phi_G(e) = \{v,w\}$ then $\phi_G(e) = \{u,w\}$, and finally returns the resulting graph.
	\end{definition}

	\begin{center}
		\input{images/contraction.tex}
	\end{center}

		The medial graph $G^\times$ of a connected plane graph $G$ is a graph such that there is a bijection between $V(G^\times)$ and $\mathbb{E}(G)$ and such that for each face $f$ of $G$, there's an edge $e^\times \in \mathbb{E}(G^\times)$ incident to a pair of vertices $u^\times,v^\times \in V(G^\times)$ if edges $u,v \in \mathbb{E}(G)$ are consecutive in $f$.
	\end{definition}

	\begin{center}
		\input{images/medial.tex}
	\end{center}

@@

		The dual graph $G^*$ of a plane graph $G$ is a graph with a bijection between the set of faces of $G$ and $V(G^*)$ and a bijection between $\mathbb{E}(G)$ and $\mathbb{E}(G^*)$ such that an edge $e \in \mathbb{E}(G)$ that separates two faces $f_1$,$f_2$ of $G$ is an edge $e^* \in \mathbb{E}(G^*)$ incident to $f_1^*$ and $f_2^*$.
	\end{definition}

	\begin{center}
		\input{images/dual.tex}
	\end{center}

	A \textit{tree} is a connected graph with no cycles.

	A \textit{leaf} $v$ of a tree $T$, is a vertex $v \in V(T)$ of degree 1.

	An \textit{internal vertex} $v$ of a tree $T$, is a vertex $v \in V(T)$ of degree at least 2.

	Let the \textit{leaf set} of a tree $T$, denoted $L(T)$, be the set of all vertices of $T$ that are also leaves.

	A \textit{Branch Decomposition} $(B_G, \delta_G)$ of a graph $G$ consists of firstly, a tree $B_G$ where every internal vertex of $B_G$ has degree 3; in other words, $B_G$ is an unrooted binary tree, and secondly a bijection $\delta_G$ between $\mathbb{E}(G)$ and $L(B_G)$.

	Removing any edge $e \in \mathbb{E}(B_G)$ partitions $B_G$ into 2 trees $P_e$ and $Q_e$. The set $L(P_e) \cap L(Q_e)$ is called a \textit{middle set} of $B_G$ given $e$, denoted $Z(B_G,e)$. The maximal cardinality of any middle set of $B_G$ given any $e$ of $B_G$ is the width of $B_G$; in other words, the width of $B_G$ is $\max\{ |Z(B, e)| \st e \in \mathbb{E}(B) \}$.

	There might exist many branch decompositions of a graph $G$.

@@

	A \textit{Carving Decomposition} $(C_G, \lambda_G)$ of a graph $G$ consists of firstly, a tree unrooted binary tree $C_G$ and secondly a bijection $\lambda_G$ between $V(G)$ and $L(C_G)$.

	Removing any edge $e \in \mathbb{E}(C_G)$ partitions $C_G$ into 2 trees $P_e$ and $Q_e$. The set $L(P_e) \cap L(Q_e)$ is called a \textit{middle set} of $C_G$ given $e$, denoted $Z(C_G,e)$. The maximal cardinality of any middle set of $C_G$ given any $e$ of $C_G$ is the width of $C_G$; in other words, the width of $C_G$ is $\max\{ |Z(C_G, e)| \st e \in \mathbb{E}(C_G) \}$.

	There might exist many carving decompositions of a graph $G$.

	A \textit{Minimal Carving Decomposition} of $G$ is any carving decomposition of $G$ of minimal width among all carving decompositions of $G$.

@@

		Input: Given a simple connected planar graph $G$.

@@

		Input: Given a simple connected graph $G$.

@@

	Problem \ref{def:mbdp} is the overarching problem, and can be broken down into many smaller subproblems.

	Considering a graph $G$, you can compute a minimal branch decomposition $(B_G, \delta_G)$ of $G$ from a minimal carving decomposition $(C_{G^\times}, \lambda_{G^\times})$ of the medial graph $G^\times$ of $G$. Do this by replacing the vertices in leaves of the carving decomposition with edges, using the mapping between edges and vertices from computing the medial graph;

	Therefore problem \ref{def:mbdp} break down into problems \ref{problem:mcd-to-mbd}, \ref{problem:medial} and \ref{problem:mcd}.

@@

	A quiet subgraph $Q(M, k, e)$, for some graph $M$, some noise level $k$ and some $e \in \mathbb{E}(M)$, is a subgraph of $M$ with $V(Q(M, k, e)) = V(M)$ and

	$\mathbb{E}(Q(M, k, e)) = \{e_1 \st $ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

	\begin{problem}\label{problem:noisy-subgraph}
		Given a plane graph $M$ that might have parallel edges, an edge $e \in \mathbb{E}(M)$, and noise level $k \in \mathbb{N}$, output the quiet subgraph $Q(M, k, e)$.
	\end{problem}

@@

	Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \st e \in \mathbb{E}(M) \land C$ is a component of $Q(M, k, e) \}$.

@@

		I assume that the input graph $G$ is a rotation system of a planar graph; an adjacency list such that the neighborhoods are given in clockwise ordering according to some plane embedding of $G$.

@@

		As the resulting graph is later given as an argument to functions assuming a rotation system of a planar graph, the implementation needs to preserve this invariant when contracting.

		As this contraction is a contraction of ALL edges connecting a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the rotation system could be difficult, but luckily in this context, it is irrelevant.

@@

		% https://en.wikipedia.org/wiki/Rotation_system
